Icicle: write once, run once

### 2016-05-26 2016fhpc/section/01-Introduction.tex
Work on bounded buffers

A joint query like this can be converted to a back-end language such as SQL, but doing it by hand it tedious and error prone. As the three sub queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter. We instead need to lift each predicate to an expression-level conditional:
\begin{code}
  SELECT COUNT(IF(open > close, 1, 0))
       , COUNT(IF(open < close, 1, 0))
       , SUM  (IF(open > close, open, 0))
       / COUNT(IF(open > close, 1, 0))
  FROM stocks;
\end{code}
As we see, the result of query fusion also tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.

Joint queries such as the stocks example can be evaluated in a streaming, incremental fashion. As a counter example, suppose we have a table with two fields @key@ and @value@, and we wish to find the mean of values whose key matches the last one in the table. We might try something like:
\begin{code}
  table kvs { key : Text; value : Real }
  query avg = let k = last key
              in  filter (key == k) of mean value;
\end{code}
Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} implied by the above query cannot be evaluated in a streaming fashion. While streaming through the table we always have access to the last key in the stream, but to perform the filter to select rows that match this key would require that we stream through the table again from the start. We need a better solution. With this in mind, the contributions of this paper are:
\begin{itemize}
\item
We present a type system that guarantees any two queries on the same input table can be fused, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};

\item
We present an fold-based intermediate language, which allows the query fusion transformation to be a simple matter of appending two programs\sref{s:IcicleCore};

\item
We present production benchmarks of our generated program, which compare favourably against standard unix utilities, and outperforms an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
\end{itemize}

>>>>

A joint query like this can be converted to a back-end language such as SQL, but doing so by hand is tedious and error prone. As the three sub queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter. We instead need to lift each predicate to an expression-level conditional:
\begin{code}
  SELECT COUNT(IF(open > close, 1, 0))
       , COUNT(IF(open < close, 1, 0))
       , SUM  (IF(open > close, open, 0))
       / COUNT(IF(open > close, 1, 0))
  FROM stocks;
\end{code}
As we see, the result of query fusion also tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.

Joint queries such as the stocks example can be evaluated in a streaming, incremental fashion, which allows us to update the result as we receive new data. As a counter example, suppose we have a table with two fields @key@ and @value@, and we wish to find the mean of values whose key matches the last one in the table. We might try something like:
\begin{code}
  table kvs { key : Date; value : Real }
  query avg = let k = last key
              in  filter (key == k) of mean value;
\end{code}
Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} implied by the above query cannot be evaluated in an inceremental fashion. While streaming through the table we always have access to the last key in the stream, but to perform the filter to select rows that match this key would require that we stream through the table again from the start. We need a better solution. With this in mind, the contributions of this paper are:
\begin{itemize}
\item
We present a domain specific language and type system that guarantees any two queries on the same input table can be fused, and allows queries to be resumed as new data is received\sref{s:IcicleSource};

\item
We present an fold-based intermediate language, which allows the query fusion transformation to be a simple matter of appending two intermediate programs\sref{s:IcicleCore};

\item
We present production benchmarks of Icicle compiled code, which compare favourably against standard unix utilities, and outperform an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
\end{itemize}

### 2016-05-26 2016fhpc/section/02-Source.tex
Work on bounded buffers

To allow incremental computation, all Icicle queries must execute in a single pass over the input stream. Of course, not all queries can be executed in a single pass: the key examples are anything that require random access indexing, or otherwise need to access data in an order that is different to what the stream will provide. However, as we saw in the introduction, although it my not be possible to evaluate a particular \emph{algorithm} in an streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm. Here is the unstreamable example from the introduction:
\begin{code}
  table kvs { key : Text; value : Real }
  query avg = let k = last key
              in  filter (key == k) of mean value;
\end{code}

The problem with this is that the value of @last key@ is only availble once we have reached the end of the stream, but the @filter@ operation needs this value to process the very first element in the same stream. We distinguish between these two access patterns by giving them different names: we say that @last key@ is an \emph{aggregate} as to compute it we need to have consumed the \emph{entire stream}, whereas the filter predicate is an \emph{element}-wise computation that only needs access to the current element read from the stream.

The trick to compute our average in an incremental fashion is to recognise the @filter@ operation is serving to select a particular sub-set of values from the input, but the value computed from this sub-set depends only on the values in that sub-set, and no other information. Instead of computing the mean of a single subset whose identity is only known at the end of the stream, we can instead compute the mean of \emph{all possible subsets}, and return the required one once we know what that is:
\begin{code}
  table kvs { key : Text; value : Real } 
  query avg = let k    = last  key in
              let avgs = group key of mean value
              in  lookup k avgs
\end{code}

@@

% ---------------------------------------------------------
\subsection{The Stage Restriction}
To ensure that Icicle queries can be evaluated in a single pass, we use a modal type system inspired by staged computation~\cite{davies2001modal}. We use two modalities, @Element@ and @Aggregate@. Values of type @Element@~$\tau$ are taken from input stream on a per-element basis, whereas values of type @Aggregate@~$\tau$ are only available once the entire stream has been consumed. In the expression @filter (key == k) of mean value@, the variable @key@ has type @Element Text@ while the variable @k@ has type @Aggregate Text@. Attempting to compile this query in Icicle will produce a type error complaining that elements cannot be compared with aggregates.

% ---------------------------------------------------------
\subsection{Finite Streams and Synchronous Data Flow}
As opposed to synchronous data flow languages such as {\sc Lustre}\cite{halbwachs1991synchronous}, all streams processed by Icicle are conceptually finite in length. Icicle is fundamentally a query language, which queries finite streams of data held in a non-volatile store. In contrast Lustre operates on conceptually infinite streams, such as those found in real-time control systems (like in airplanes). In Icicle, the ``last'' element in a stream is the last one that was in the file read from disk. In Lustre, the ``last'' element in a stream is the one that was most recently received. If the ``unstreamable'' version of the query from \S2 was converted into Lustre then it would execute, but the filter predicate would compare the last key with the most recent key from the stream, which is the key itself. This means the filter predicate would always be true, and the query would return the mean of the entire stream. Applying the Icicle type system to our queries imposes the natural stage restriction associated with finite streams, so there are distinct ``during'' (element) and ``after'' (aggregate) stages.

@@

\TODO{Point out that query from S2 can be evaluated incrementally}

The modalities are automatically boxed and unboxed, so that if a function expecting a pure value is passed an @Element@ computation, the result of the function becomes an @Element@ computation too. For example, if @open : Element Int@, then @open == 1@ has type @Element Bool@. @Element@ and @Aggregate@ are mutually exclusive: a computation cannot be both an element and an aggregate.

@@

% ---------------------------------------------------------
\subsection{Bounded buffer restriction}
\label{s:IcicleSource:bounded}
\TODO{Point out that the group uses space proportional to the number of keys in the input, which may be as large as the number of elements in the stream}

As Icicle is designed to be a streaming language, the amount of data to be streamed may not necessarily fit in memory.
Any operation which requires a buffer must be bounded in size, as an unbounded buffer could potentially grow too large to fit in memory.

Here is an example stream program that requires unbounded buffering:
it takes an input stream @xs@, and filters it into those above zero, and those below zero.
These two filtered streams are then joined together pairwise, so the first positive element is paired with the first negative element, and so on.
The program, written in Haskell:

\begin{code}
zip (filter (>0) xs) (filter (<0) xs)
\end{code}

This program requires an unbounded buffer: if the input stream contains ten positive values followed by one negative value, all ten of the positive values must be held onto until the negative value is seen.
Similarly, if the entire stream is positive, all of the elements must be retained until the very end, just in case a negative value shows up.

In Icicle, we outlaw this kind of program by implicitly threading the input stream through operations.
The streams themselves are not materialised in the program: stream operations like @fold@ and @filter@ do not take parameters of the streams, but instead operate on the context stream.
By removing the explicit stream parameter, stream elements can only be joined from the same context, when both elements are available.

For example, in the query @filter p in mean value@, the @mean value@ is only applied to stream values which satisfy the predicate @p@.

This is a different approach than existing synchronous streaming languages\cite{mandel2010lucy} and flow fusion\cite{lippmeier2013data}, which perform clock analysis or `rate inference'.
Here, streams are given clock types denoting when the streams have elements available, and only streams with the same clock can be zipped together.
Filters produce a stream with a different output clock to its input, so different filters cannot be zipped together.
Our system requires no clock analysis, but is less expressive as streams are not `first class'.

>>>>

To allow incremental computation, all Icicle queries must execute in a single pass over the input stream. Of course, not all queries can be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in an order that is different to what a stream can provide. However, as we saw in the introduction, although it my not be possible to evaluate a particular \emph{algorithm} in an streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm. Here is the unstreamable example again from the introduction:
\begin{code}
  table kvs { key : Date; value : Real }
  query avg = let k = last key
              in  filter (key == k) of mean value;
\end{code}

The problem with this is that the value of @last key@ is only availble once we have reached the end of the stream, but the @filter@ operation needs it to process the very first element in the same stream. We distinguish between these two access patterns by giving them different names: we say that @last key@ is an \emph{aggregate}, because to compute it we need to have consumed the \emph{entire stream}, whereas the filter predicate is an \emph{element}-wise computation because it only needs access to the current element in the stream.

The trick to compute our average in an incremental fashion is to recognise that the @filter@ operation is selecting a particular sub-set of values from the input, but the value computed from this sub-set depends only on the values in that sub-set, and no other information. Instead of computing the mean of a single subset whose identity is only known at the end of the stream, we can instead compute the mean of \emph{all possible subsets}, and return the required one once we know what that is:
\begin{code}
  table kvs { key : Date; value : Real } 
  query avg = let k    = last  key in
              let avgs = group key of mean value
              in  lookup k avgs
\end{code}

@@

% ---------------------------------------------------------
\subsection{The Stage Restriction}
To ensure that Icicle queries can be evaluated in a single pass, we use a modal type system inspired by staged computation~\cite{davies2001modal}. We use two modalities, @Element@ and @Aggregate@. Values of type @Element@~$\tau$ are taken from input stream on a per-element basis, whereas values of type @Aggregate@~$\tau$ are only available once the entire stream has been consumed. In the expression @(filter (key == k) of mean value)@, the variable @key@ has type @Element Text@ while @k@ has type @Aggregate Text@. Attempting to compile this query in Icicle will produce a type error complaining that elements cannot be compared with aggregates.

Note that the types of pure values such as constants are automatically promoted to have the required modality, for example, if we have @open == 1@ and @open : Element Int@ then the constant @1@ is automatically promoted to have type @Element Int@ as well.

% ---------------------------------------------------------
\subsection{Finite Streams and Synchronous Data Flow}
As opposed to synchronous data flow languages such as {\sc Lustre}~\cite{halbwachs1991synchronous}, all streams processed by Icicle are conceptually finite in length. Icicle is fundamentally a query language, which queries finite tables of data held in a non-volatile store, but does so in a streaming manner. In contrast, Lustre operates on conceptually infinite streams, such as those found in real-time control systems (like to fly airplanes). In Icicle, the ``last'' element in a stream is the last one that appears in the file on disk. In Lustre, the ``last'' element in a stream is the one that was most recently received. If the ``unstreamable'' version of the query from \S2 was converted into Lustre syntax then it would execute, but the filter predicate would compare the last key with the most recent key from the stream, which is the key itself. This means the filter predicate would always be true, and the query would return the mean of the entire stream. Applying the Icicle type system to our queries imposes the natural stage restriction associated with finite streams, so there are distinct ``during'' (element) and ``after'' (aggregate) stages.

@@

Suppose we apply a query large table and record the result. Tomorrow morning we receive more data and add it to the table. We would like to recompute the result of this query without needing to process all data from the start of the table. We can do this by remembering the values of all intermediate aggregates that are computed in the query, and updating them as new data arrives. In the @avg@ example from \S2 these aggregates are @k@ and @avgs@.

@@

% ---------------------------------------------------------
\subsection{Bounded buffer restriction}
\label{s:IcicleSource:bounded}
Icicle queries are intended to process tables of arbitrary size that may not fit in memory. Due to this, each query must execute without requiring buffer space proportional to the size of the input. For example, here is a simple Haskell expression which cannot be executed without reserving a buffer at least the same size as the input:
\begin{code}
      zip (filter (> 0) xs) (filter (< 0) xs)
\end{code}

Suppose @(xs :: Stream Int)@. The expression takes the input stream @xs@ and selects the elements greater than zero, and pairs them with elements that are less than zero. This expression requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then the positive values must be retained before any output can be produced.

The language constructs of Icicle are designed so that queries that would require unbounded buffering cannot be written, with a one major cavet that we will discuss in a moment. In Icicle, the stream being processed (such as @xs@ above) is implicit in the defined query, and is never named explicitly. Constructs such as @filter@ and @fold@ do not take the name of the stream as argument, but instead operate on stream defined in the context. In all the example Icicle queries so far, the name of the table / stream is not mentioned in the body of the query, and serves as a comment only. Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query. In the expression @filter p in mean value@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.

Finally, note that our major caveat is that @group@ combinator we used in \S2 uses space proportional to the number of distinct \emph{keys} in the input stream. For our applications the keys are commonly company names, customer names, and days of the year. Our production system knows that these types are bounded in size and that maps from keys to values will fit easily in memory. Attempting to group by values of a type with an unbounded number of members, such as a @Real@ or @String@ results in a compile-time warning.

% Here, streams are given clock types denoting when the streams have elements available, and only streams with the same clock can be zipped together. Filters produce a stream with a different output clock to its input, so different filters cannot be zipped together. Our system requires no clock analysis, but is less expressive as streams are not `first class'.

### 2016-05-26 2016fhpc/section/XX-Related.tex
ok
Evaluation rules
Work on bounded buffers

@@

>>>>

\TODO{fold this into related}
The Icicle approach to bounded buffering is simpler but less expressive than both existing synchronous data flow languages such as Lucy~\cite{mandel2010lucy}, as well as our prior work on flow fusion~\cite{lippmeier2013data}. These systems perform a clock, or \emph{rate} analysis of the streams being processed. Stream combinators with multiple inputs, such as @zip@, are assigned types that require their stream arguments to have the same rate --- meaning that elements always arrive in lockstep and the combinators themselves do not need to perform their own buffering.

@@

Even better is Nie and Cheng's approach\cite{nie2007efficient}, which does not seem to be more complete or lower complexity than Gulwani and Necula, just lower constant factors.
